Defining Transcriptome using RNA Sequencing\\ of {\em Streptococcus mutans}

### 2012-04-12 package/manuscript.Rnw
Just an update.

@@

\subsection*{Short reads alignments}
About 20 millions of short reads were sequenced for each sample.  We used a
short read aligner, \program{BWA} \citep{Li2009a}, to map the sequenced short
reads on the reference genome, \textit{S.\ mutans} UA159 (NCBI accession
NC\_004350). Because the aligner allowed a few gaps for efficient alignments of
millions of sequences of size being 100 base pairs long, short reads that
consisted of RNAs with sequencing adapter sequences would not be mapped.  We
used \program{cutadapt} \citep{Martin2011} to remove parts of adapter sequences
from short reads before mapping on the reference genome.  Mapped short read
alignments were converted into readable formats using \program{samtools}
\citep{Li2009b}.

@@

<<echo=FALSE,result=hide>>=
notExpressedTx <- smutansData.txGrCombined[elementMetadata(smutansData.txGrCombined)$zero > 0.1 & elementMetadata(smutansData.txGrCombined)$pileup < 5]
grTx <- smutansData.txGrCombined[!(elementMetadata(smutansData.txGrCombined)$zero > 0.1 & elementMetadata(smutansData.txGrCombined)$pileup < 5)]
# We query genes and annotations (genes and RNAs) againt tx's.
matchGenesTx <- findOverlaps(smutansData.txGenes,grTx)
matchAnnotationTx <- findOverlaps(smutansData.txAnnotation,grTx)
# I use gaps with IRanges because gaps with GenomicRanges does not work as expected.
grNoAnnotation <- gaps(ranges(smutansData.txAnnotation))
grNoncoding <- gaps(ranges(smutansData.txGenes))
grRNAz <- ranges(smutansData.txRNAz)
# 
matchUnannotatedRNA <- findOverlaps(grRNAz,grNoAnnotation,type="within")
txRNAz.unannotated <- grRNAz[matchMatrix(matchUnannotatedRNA)[,1]]

@@

# Expressed transcripts without annotated genes
grTxWGenes <- grTx[unique(matchMatrix(matchGenesTx)[,2])]
grTxWOGenes <- grTx[-unique(matchMatrix(matchGenesTx)[,2])]
grTxWOAnnotation <- grTx[-unique(matchMatrix(matchAnnotationTx)[,2])]

# Find transcripts with genes with conflicted strands
grGenesPlus <- smutansData.txGenes[strand(smutansData.txGenes)=='+']
grGenesMnus <- smutansData.txGenes[strand(smutansData.txGenes)=='-']
#mtchPlus <- findOverlaps(grGenesPlus,grTxWGenes)
#mtchMnus <- findOverlaps(grGenesMnus,grTxWGenes)
#txWithConflictedGenes <- intersect(matchMatrix(mtchPlus)[,2],matchMatrix(mtchMnus)[,2])

@@

# Find the number of genes in segments.
matchGenesTx2 <- findOverlaps(smutansData.txGenes,grTx2,type="within")
number.genes.tx <- runLength(Rle(matchMatrix(matchGenesTx2)[,"subject"]))

# Length distribution of UTR regions
# grTx2 <- grTxWGenes[-txWithConflictedGenes] 
# Lengths of 5' and 3' UTRs
mtchPlus <- findOverlaps(grGenesPlus,grTx2,type="within")
mtchMnus <- findOverlaps(grGenesMnus,grTx2,type="within")
length5UTR <- c()
length3UTR <- c()
for (i in unique(matchMatrix(mtchPlus)[,"subject"])) {
  x <- matchMatrix(mtchPlus)[matchMatrix(mtchPlus)[,"subject"]==i,"query"]
  y <- min(start(grGenesPlus[x])) - start(grTx2[i])
  z <- end(grTx2[i]) - max(end(grGenesPlus[x])) 
  length5UTR <- c(length5UTR,y)
  length3UTR <- c(length3UTR,z)
}
for (i in unique(matchMatrix(mtchMnus)[,"subject"])) {
  x <- matchMatrix(mtchMnus)[matchMatrix(mtchMnus)[,"subject"]==i,"query"]
  y <- min(start(grGenesMnus[x])) - start(grTx2[i])
  z <- end(grTx2[i]) - max(end(grGenesMnus[x])) 
  length5UTR <- c(length5UTR,z)
  length3UTR <- c(length3UTR,y)
}

# Correlation of expression levels of adjacent genes.
pileup.x <- lapply(ranges(smutansData.txGenes), function(x) seqselect(smutansData.txPileup, x))
pileup.y <- mapply(function(x,y) sum(x)/y, x=pileup.x, y=width(smutansData.txGenes))
elementMetadata(smutansData.txGenes)["score"] <- pileup.y
mtch3 <- findOverlaps(smutansData.txGenes,grTx2)
x <- rle(matchMatrix(mtch3)[,"subject"])
comp.exp.two.genes <- c()
for (i in x$values[x$lengths > 1]) {
  y <- combn(matchMatrix(mtch3)[,"query"][matchMatrix(mtch3)[,"subject"]==i],2)
  for (j in 1:dim(y)[2]) {
    comp.exp.two.genes <- cbind(comp.exp.two.genes,as.matrix(elementMetadata(smutansData.txGenes)["score"][y[,j],]))
  }
}
# Correlation of expression levels of adjacent genes between transcripts.
comp.exp.two.genes.between.transcript <- c()
for (i in 1:length(x$values)) {
  if (i == length(x$values)) {
    break
  }
  y1 <- matchMatrix(mtch3)[,"query"][matchMatrix(mtch3)[,"subject"]==x$values[i]]
  i2 <- i + 1
  y2 <- matchMatrix(mtch3)[,"query"][matchMatrix(mtch3)[,"subject"]==x$values[i2]]
  l <- as.matrix(do.call(expand.grid, list(y1,y2)))
  for (j in 1:dim(l)[1]) {
    comp.exp.two.genes.between.transcript <- cbind(comp.exp.two.genes.between.transcript,as.matrix(elementMetadata(smutansData.txGenes)["score"][l[j,],]))
  }
}
@

@@

We obtained a set of segments of RNA-seq expressions for each of the 13 RNA-seq
samples.  All of the segments for each set of RNA-seq expression segments were
scored by the average proportions of segments shared by the segments in the rest
of sets.  We placed more highly ranked segments first on the reference genome
based on the scores, and purged more lowly ranked segments from the
transcriptome map. In doing so, the
final set of segments was formed by placing non-overlapping highly ranked
segments.  Average site-wise expression levels of the segments 
are shown at figure \ref{figGrCombinedExpression}. 
We designated segments as expressed that had site-wise pileup values
greater than 5 and proportion of sites of zero site-wise pileup values less than
10\%.  A gene was also designated as an expressed if the gene belonged to an
expressed segment.  Some of the predicted segments were not expressed or very
lowly expressed (\Sexpr{length(notExpressedTx)} 
out of 
\Sexpr{length(smutansData.txGrCombined)}).
Of the total \Sexpr{length(smutansData.txGenes)} of genes,
\Sexpr{length(matchMatrix(matchGenesTx)[,1])}
(\Sexpr{as.integer(length(matchMatrix(matchGenesTx)[,1])/length(smutansData.txGenes)*100)}\%) 
were designated as expressed. 
Genes that were not designated as expressed could be due to false gene
annotations or possibility of true negative expression under the experimental
conditions. 
Figure \ref{figLength} shows the length distribution of expressed and
not-expressed genes. 
Of the \Sexpr{length(grTx)} expressed segments, 
\Sexpr{length(grTxWOAnnotation)} contained no annotated genes or RNAs.
Many of the segments might cover gaps between real operons.
\fixme{Examine these unannotated regions in the UCSC genome tracks.}

Figure \ref{figTranscriptLength} shows the length distribution of expressed
segments with annotated coding genes and unannotated regions.  Unannotated
regions could include potential non-coding regions and novel coding genes.
Segments of unannotated regions appeared to be shorter than those with
coding genes. Most of the unannotated regions with length being near 360 bp
could be non-coding regions, and the rest of the unannotated regions with length
being near 1000 bp could be new coding genes. Some of the unannotated genes
could be pseudogene regions.  We used strandnesses of known genes to find
strands of segments although we could not tell strands of the predicted
transcripts solely based on our RNA-seq data because of our strand-nonspecific
RNA-seq technique.  Of
the \Sexpr{length(grTx)} expressed transcripts, 
\Sexpr{sum(strand(grTx)=='*')}
included annotated genes of both strands
(\Sexpr{as.integer(sum(strand(grTx)=='*')/length(grTx)*100)}\%).
We filtered out the transcripts with genes with conflicted both strands leaving
\Sexpr{length(grTx2)} expressed transcripts. 
Figure \ref{figUTRLength} shows length distributions of 5' and 3' UTR of the
expressed transcripts without genes with conflicted strands.  Many transcripts
appeared to start and end near the boundaries of annotated genes.  Gene
expression levels for adjacent genes within expressed transcripts were more
correlated than those between adjacent transcripts 
(fig.\ \ref{figExpressionTwoGene}).
Among the \Sexpr{length(unique(matchMatrix(matchGenesTx2)[,"subject"]))} segments
with genes of all of the same strands, 
\Sexpr{sum(number.genes.tx==1)} segments contained a single gene while 
\Sexpr{sum(number.genes.tx > 1)} contained more than one gene.

>>>>

\section*{Main findings}
The resolution of RNA-Seq has advanced many aspects of the existing gene
annotation, including the determination of gene boundaries for known genes and
the identification of novel transcribed regions in many of the bacteria that
have previously been sequenced.
For example, for L. monocytogenes, 517 primary operons were identified that
encode 1719 genes, which represent 60% of the annotated genome
The knowledge gained regarding untranslated regulatory regions, operons and
sRNAs is triggering re-evaluation and re-annotation of many bacterial genomes.

@@

\subsection*{Short reads alignments}
About 20 millions of short reads were sequenced for each sample 
(table \ref{count}).  We used a
short read aligner, \program{BWA} \citep{Li2009a}, to map the sequenced short
reads on the reference genome, \textit{S.\ mutans} UA159 (NCBI accession
NC\_004350). Because the aligner allowed a few gaps for efficient alignments of
millions of sequences of size being 100 base pairs long, short reads that
consisted of RNAs with sequencing adapter sequences would not be mapped.  We
used \program{cutadapt} \citep{Martin2011} to remove parts of adapter sequences
from short reads before mapping on the reference genome.  Mapped short read
alignments were converted into readable formats using \program{samtools}
\citep{Li2009b}.

@@

<<echo=FALSE,result=hide>>=
notExpressedTx <- smutansData.txGrCombined[elementMetadata(smutansData.txGrCombined)$zero > 0.1 & elementMetadata(smutansData.txGrCombined)$pileup < 5]
grTx <- smutansData.txGrCombined[!(elementMetadata(smutansData.txGrCombined)$zero > 0.1 & elementMetadata(smutansData.txGrCombined)$pileup < 5)]
# We query genes and annotations (genes and RNAs) againt tx's.
matchGenesTx <- findOverlaps(smutansData.txGenes,grTx)
matchAnnotationTx <- findOverlaps(smutansData.txAnnotation,grTx)
# I use gaps with IRanges because gaps with GenomicRanges does not work as expected.
grNoAnnotation <- gaps(ranges(smutansData.txAnnotation))
grNoncoding <- gaps(ranges(smutansData.txGenes))
grRNAz <- ranges(smutansData.txRNAz)
# 
matchUnannotatedRNA <- findOverlaps(grRNAz,grNoAnnotation,type="within")
txRNAz.unannotated <- grRNAz[as.matrix(matchUnannotatedRNA)[,1]]

@@

# Expressed transcripts without annotated genes
grTxWGenes <- grTx[unique(as.matrix(matchGenesTx)[,2])]
grTxWOGenes <- grTx[-unique(as.matrix(matchGenesTx)[,2])]
grTxWOAnnotation <- grTx[-unique(as.matrix(matchAnnotationTx)[,2])]

# Find transcripts with genes with conflicted strands
grGenesPlus <- smutansData.txGenes[strand(smutansData.txGenes)=='+']
grGenesMnus <- smutansData.txGenes[strand(smutansData.txGenes)=='-']
#mtchPlus <- findOverlaps(grGenesPlus,grTxWGenes)
#mtchMnus <- findOverlaps(grGenesMnus,grTxWGenes)
#txWithConflictedGenes <- intersect(as.matrix(mtchPlus)[,2],as.matrix(mtchMnus)[,2])

@@

# Find the number of genes in segments.
matchGenesTx2 <- findOverlaps(smutansData.txGenes,grTx2,type="within")
number.genes.tx <- runLength(Rle(as.matrix(matchGenesTx2)[,"subjectHits"]))

# Length distribution of UTR regions
# grTx2 <- grTxWGenes[-txWithConflictedGenes] 
# Lengths of 5' and 3' UTRs
mtchPlus <- findOverlaps(grGenesPlus,grTx2,type="within")
mtchMnus <- findOverlaps(grGenesMnus,grTx2,type="within")
length5UTR <- c()
length3UTR <- c()
for (i in unique(as.matrix(mtchPlus)[,"subjectHits"])) {
  x <- as.matrix(mtchPlus)[as.matrix(mtchPlus)[,"subjectHits"]==i,"queryHits"]
  y <- min(start(grGenesPlus[x])) - start(grTx2[i])
  z <- end(grTx2[i]) - max(end(grGenesPlus[x])) 
  length5UTR <- c(length5UTR,y)
  length3UTR <- c(length3UTR,z)
}
for (i in unique(as.matrix(mtchMnus)[,"subjectHits"])) {
  x <- as.matrix(mtchMnus)[as.matrix(mtchMnus)[,"subjectHits"]==i,"queryHits"]
  y <- min(start(grGenesMnus[x])) - start(grTx2[i])
  z <- end(grTx2[i]) - max(end(grGenesMnus[x])) 
  length5UTR <- c(length5UTR,z)
  length3UTR <- c(length3UTR,y)
}

# Correlation of expression levels of adjacent genes.
pileup.x <- lapply(ranges(smutansData.txGenes), function(x) seqselect(smutansData.txPileup, x))
pileup.y <- mapply(function(x,y) sum(x)/y, x=pileup.x, y=width(smutansData.txGenes))
elementMetadata(smutansData.txGenes)["score"] <- pileup.y
mtch3 <- findOverlaps(smutansData.txGenes,grTx2)
x <- rle(as.matrix(mtch3)[,"subjectHits"])
comp.exp.two.genes <- c()
for (i in x$values[x$lengths > 1]) {
  y <- combn(as.matrix(mtch3)[,"queryHits"][as.matrix(mtch3)[,"subjectHits"]==i],2)
  for (j in 1:dim(y)[2]) {
    comp.exp.two.genes <- cbind(comp.exp.two.genes,as.matrix(elementMetadata(smutansData.txGenes)["score"][y[,j],]))
  }
}
# Correlation of expression levels of adjacent genes between transcripts.
comp.exp.two.genes.between.transcript <- c()
for (i in 1:length(x$values)) {
  if (i == length(x$values)) {
    break
  }
  y1 <- as.matrix(mtch3)[,"queryHits"][as.matrix(mtch3)[,"subjectHits"]==x$values[i]]
  i2 <- i + 1
  y2 <- as.matrix(mtch3)[,"queryHits"][as.matrix(mtch3)[,"subjectHits"]==x$values[i2]]
  l <- as.matrix(do.call(expand.grid, list(y1,y2)))
  for (j in 1:dim(l)[1]) {
    comp.exp.two.genes.between.transcript <- cbind(comp.exp.two.genes.between.transcript,as.matrix(elementMetadata(smutansData.txGenes)["score"][l[j,],]))
  }
}
@

@@

We obtained a set of segments of RNA-seq expressions for each of the 13 RNA-seq
samples.  All of the segments for each set of RNA-seq expression segments were
scored by the average proportions of segments shared by the segments in the rest
of sets.  We placed more highly ranked segments first on the reference genome
based on the scores, and purged more lowly ranked segments from the
transcriptome map. In doing so, the
final set of segments was formed by placing non-overlapping highly ranked
segments.  Average site-wise expression levels of the segments 
are shown at figure \ref{figGrCombinedExpression}. 
We designated segments as expressed that had site-wise pileup values
greater than 5 and proportion of sites of zero site-wise pileup values less than
10\%.  A gene was also designated as an expressed if the gene belonged to an
expressed segment.  Some of the predicted segments were not expressed or very
lowly expressed (\Sexpr{length(notExpressedTx)} 
out of 
\Sexpr{length(smutansData.txGrCombined)}).
Of the total \Sexpr{length(smutansData.txGenes)} of genes,
\Sexpr{length(as.matrix(matchGenesTx)[,1])}
(\Sexpr{as.integer(length(as.matrix(matchGenesTx)[,1])/length(smutansData.txGenes)*100)}\%) 
were designated as expressed. 
Genes that were not designated as expressed could be due to false gene
annotations or possibility of true negative expression under the experimental
conditions. 
Figure \ref{figLength} shows the length distribution of expressed and
not-expressed genes. 
Of the \Sexpr{length(grTx)} expressed segments, 
\Sexpr{length(grTxWOAnnotation)} contained no annotated genes or RNAs.
Many of the segments might cover gaps between real operons.
\fixme{Examine these unannotated regions in the UCSC genome tracks.}

Figure \ref{figTranscriptLength} shows the length distribution of expressed
segments with annotated coding genes and unannotated regions.  Unannotated
regions could include potential non-coding regions and novel coding genes.
Segments of unannotated regions appeared to be shorter than those with
coding genes. Most of the unannotated regions with length being near 360 bp
could be non-coding regions, and the rest of the unannotated regions with length
being near 1000 bp could be new coding genes. Some of the unannotated genes
could be pseudogene regions.  We used strandnesses of known genes to find
strands of segments although we could not tell strands of the predicted
transcripts solely based on our RNA-seq data because of our strand-nonspecific
RNA-seq technique.  Of
the \Sexpr{length(grTx)} expressed transcripts, 
\Sexpr{sum(strand(grTx)=='*')}
included annotated genes of both strands
(\Sexpr{as.integer(sum(strand(grTx)=='*')/length(grTx)*100)}\%).
We filtered out the transcripts with genes with conflicted both strands leaving
\Sexpr{length(grTx2)} expressed transcripts. 
Figure \ref{figUTRLength} shows length distributions of 5' and 3' UTR of the
expressed transcripts without genes with conflicted strands.  Many transcripts
appeared to start and end near the boundaries of annotated genes.  Gene
expression levels for adjacent genes within expressed transcripts were more
correlated than those between adjacent transcripts 
(fig.\ \ref{figExpressionTwoGene}).
Among the \Sexpr{length(unique(as.matrix(matchGenesTx2)[,"subjectHits"]))} segments
with genes of all of the same strands, 
\Sexpr{sum(number.genes.tx==1)} segments contained a single gene while 
\Sexpr{sum(number.genes.tx > 1)} contained more than one gene.
